### 2018-06-27 latex/manuscript.tex
Minor changes and library update
Added supplementary figure 3 and corresponding figure legend
Included changes by Ines

@@

\section{Summary}

NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam). In eukaryotes, two different Nam salvage pathways exist. Phylogenetic analysis of the pathway show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates. Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT). The physiological relevance of NNMT is still enigmatic. Why is there an enzyme that removes Nam from recycling, seemingly not having any other physiological role? Despite the occurance of this Nam degrading enzyme NNMT,  we observe  a diversification of NAD dependent signalling enzymes in vertebrates. This seems counterintuitive.

@@

\section{Keywords}

>>>>

\noindent
\textit{Alternative title:} Functional interdependencies as driving force for the evolution of NAD-dependent signalling and biosynthesis

\vspace{5mm}

@@

\section*{Summary}

NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam). In eukaryotes, two different Nam salvage pathways exist. Phylogenetic analysis of the pathway show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates. Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT). The physiological relevance of NNMT is still enigmatic. Why is there an enzyme that removes Nam from recycling, seemingly not having any other physiological role? Despite the occurrence of this Nam degrading enzyme NNMT, we observe a diversification of NAD dependent signalling enzymes in vertebrates. This seems counterintuitive.

@@

\section*{Keywords}

### 2018-06-27 latex/methods.tex
Minor changes and library update
Included changes by Ines

Kinetic parameters (substrate affinity ($K_{M}$) and turnover rates ($k_{cat}$), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions. Parameter values from mammals were used if available. For enzymes not present in mammals, values from yeast were used. The full list of kinetic parameters including reference to original literature can be found in Supplementary table 1. For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}. Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively. All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}. The model will be available at the Biomodels database accession number xxx\todo[author=Ines]{Insert number}. Related figures were generated using gnuplot\todo[author=Ines]{Version?}.

@@

NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr). A list of functionally verified enzymes used as templates is given in supplementary table 2. This table also includes the length cut-offs for identified enzymes. The e-value cut-off was 1e-30 for all enzymes. Blastp parameters were set to yield maximum 20\,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively. Low-complexity filtering was disabled. Obvious sequence contaminations were removed by manual inspection of the results. The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI. Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python, using the ETE3 toolkit (Huerta-Cepas, 2010).

>>>>

Kinetic parameters (substrate affinity ($K_{M}$) and turnover rates ($k_{cat}$), substrate and product inhibitions) were retrieved from the enzyme database BRENDA and additionally evaluated by checking the original literature especially with respect to measurement conditions. Parameter values from mammals were used if available. For enzymes not present in mammals, values from yeast were used. The full list of kinetic parameters including reference to original literature can be found in supplementary table~S2. For NMNAT, the previously developed rate law for substrate competition was used \cite{Schauble2013}. Despite these modifications, Henri-Michaelis-Menten kinetics were used for all reactions except the import and efflux of Nam, which were simulated using constant flux and mass action kinetics, respectively. All simulations were performed using the steady state calculation and parameter scan options provided by COPASI 4.22 \cite{Hoops2006}. The model will be available at the Biomodels database upon publication. Related figures were generated using Gnuplot~5.0.

@@

NADA, NamPRT, and NNMT enzymes or enzyme candidates were identified using Blastp with known enzymes against the NCBI non-redundant protein sequence database (nr). A list of functionally verified enzymes used as templates is given in supplementary table~S1. This table also includes the length cut-offs for identified enzymes. The e-value cut-off was 1e-30 for all enzymes. Blastp parameters were set to yield maximum 20\,000 target sequences, using the BLOSUM62 matrix with a word size of 6 and gap opening and extension costs of 11 and 1, respectively. Low-complexity filtering was disabled. Obvious sequence contaminations were removed by manual inspection of the results. The taxonomy IDs of the species for each enzyme was derived from the accession2taxonomy database provided by NCBI. Scripts for creating, analysing, and visualising the phylogenetic tree were written in Python, using the ETE3 toolkit (Huerta-Cepas, 2010).

### 2018-06-27 latex/results.tex
Added supplementary figure 3 and corresponding figure legend
Included changes by Ines

NAD can be synthesised using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see fig.~\ref{fig:pathway_overview}). Nam and NA are together known as vitamin B3 or niacin. Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}. Due to the high turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD\todo{source}, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursors of human NAD biosynthesis.

@@

Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only NamPRT replaces NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling. NNMT seems to have arisen \textit{de novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi genes named NNMT can be found in databases. These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes. The yeast protein has later been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}. Nematodes are the only organisms that encode NNMT together with NADA without NamPRT being present. In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds. The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (fig.~S2). The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine. There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database.

@@

So why does the diversity of NAD-dependent signalling increase? And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi? Given the complexity of the NAD metabolic network, these questions are difficult to be comprehensively addressed experimentally. We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details, see experimental procedures and supplementary table~S2).

@@

The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria. To analyse this question we built a two compartment model that has a shared Nam source (for details, see experimental procedures and supplementary table~S2). One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT. Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (fig.~\ref{fig:NNMT_comp_advantage}A), but is able to maintain much higher steady state NAD concentrations at low cell division rates (fig.~\ref{fig:NNMT_comp_advantage}B). At high cell division rates, steady state concentrations in both compartments are similar. The latter might explain why in bacteria that have relatively high growth rates both systems co-exist.

>>>>

NAD can be synthesised using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see fig.~\ref{fig:pathway_overview}). Nam and NA are together known as vitamin B3 or niacin. Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}. Due to the high turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD\todo{Needs a reference}, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursors of human NAD biosynthesis.

@@

Analysing the phylogeny of the NAD recycling enzymes in Metazoa in more detail reveals that not only NamPRT replaces NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling. NNMT seems to have arisen \textit{de novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi genes named NNMT can be found in databases. These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes. The yeast protein has later been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}. Nematodes are the only organisms that encode NNMT together with NADA without NamPRT being present. In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds. The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (suppl. fig.~S2). The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine. There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database.

@@

So why does the diversity of NAD-dependent signalling increase? And why does NADA disappear in Deuterostomia although it is the predominant pathway in bacteria, plants and fungi? Given the complexity of the NAD metabolic network, these questions are difficult to be comprehensively addressed experimentally. We therefore built a dynamic model of NAD metabolism using existing kinetic data from the literature (for details, see experimental procedures and suppl. tab.~S2).

@@

The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria. To analyse this question we built a two compartment model that has a shared Nam source (for details, see experimental procedures and suppl. tab.~S2). One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT. Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (fig.~\ref{fig:NNMT_comp_advantage}A), but is able to maintain much higher steady state NAD concentrations at low cell division rates (fig.~\ref{fig:NNMT_comp_advantage}B). At high cell division rates, steady state concentrations in both compartments are similar. The latter might explain why in bacteria that have relatively high growth rates both systems co-exist.
